`SimSurvey`: an `R` package for simulation testing the design and analysis of fisheries-independent trawl surveys

### 2019-05-31 analysis/paper/sim_survey_paper.Rmd
Revise abstrat, complete draft

# Abstract

A central question to fish stock assessments is: how many samples are needed to adequately represent the population being monitored? For assessments based on trawl surveys, our ability to estimate population characteristics, such as abundance at age, is affected by multiple levels of sampling (i.e. the number of sets, lengths and ages sampled in a survey). We therefore need to assess the impact of haul design and length and age sampling on survey accuracy to determine optimal sampling effort. In this study we simulated repeated stratified-random samples from a spatially correlated field of simulated fish. These samples were then analyzed and estimates of abundance at age were compared against a known truth. By varying the sampling protocol, we demonstrate that increasing the number of sets conducted per stratum provides the greatest gains in precision when sampling a clustered population. Increasing length and age sampling efforts also improved abundance estimates, but to a lesser extent and improvements were typically marginal, and sometimes squandered, when sub-sampling effort was increased. These results indicate that high levels of length and age sampling effort can be scaled back without significant losses to precision and, in some cases, estimates of abundance at age may be improved by reducing sub-sampling effort. Time saved by reducing said sampling may afford more time to conduct additional sets and/or increase sampling efforts of other species. While this simulation is tailored to a specific case, the method may be adapted to different systems and survey procedures.

**Keywords** - length and age distribution; simulation; spatial correlation; stratified analysis; survey design

@@

It is well known that the accuracy and reliability of results from any population monitoring program hinges on good survey design and analysis [@thomas2010]. Ideally, the design and analysis of large and costly survey should be simulation tested before embarking on expensive field programs. This is especially true for the large-scale and costly fisheries-independent trawl surveys conducted by numerous regional fisheries management organizations. These surveys have become a mainstay in the management of dynamic fish stocks as they provide indices of population abundance as well as estimates of various population characteristics such as length and age frequencies [@Pennington1998]. This information forms the basis of of many stock assessments throughout the world. Efficient, effective and informed management decisions therefore require surveys and analyses that maximize information while minimizing the expense of data collection. While simulations provide a platform for exploring solutions to this optimization problem, building the necessary simulation framework is not a trivial task given the multi-stage nature of the typical sampling program and the complexity of the population processes we aim to represent. The complexity of this problem is perhaps why simulations that test the full sampling and analytical pathway of fisheries-independent surveys are rare [but see @puerta2019; @schnute2003].

Here we document **`SimSurvey`**, an **`R`** package designed to facilitate realistic simulations of fisheries-independent trawl surveys. This simulation framework has similarities to those presented by Schnute and Haigh [-@schnute2003] and Puerta et al. [-@puerta2019], however, efforts were focused here on developing a series of general and accessible functions to simplify the process of testing multiple sampling scenarios and analytical pathways. The steps taken to simulate surveys of spatial, age-structure populations are outlined below, with specific emphasis of a case study that motivated this work.

@@

The **`SimSurvey`** package was written in the programming language **`R`** [@R] and, in short, the package allows for the simulation of random or stratified-random surveys of an age-structured population that varies in space and time. **`SimSurvey`** relies heavily on functions from the **`data.table`** [@dowle2017], **`raster`** [@hijmans2016] and **`plotly`** [@sievert2018] packages for their efficient data processing, geographic and plotting facilities, respectively. All the source **`R`** code behind **`SimSurvey`** is available on GitHub (https://github.com/PaulRegular/SimSurvey) and the package can be installed via GitHub:

@@

```{r grid, fig.cap = "- Plots produced by `plot_grid` when supplied an raster object produced by `make_grid` a) using default case study settings and b) settings generates a grid with four divisions and a more linear depth gradient. In these plots, the color gradient represents depth, the grey lines deliniate divisions and white lines deliniate strata."}

@@

As noted above, the amount of RAM a user's computer has may limit the utility of the `sim_survey` function for running thousands of simulations of the same survey. The `sim_survey_parallel` function was therefore constructed to facilitate the running of thousands of simulations without running into RAM limitations. This function is set-up to run multiple `sim_survey` calls in parallel using the doParallel package [@weston2015] and, as such, multiple loops can be run using the `n_loops` argument and within each loop multiple simulations can be run (controlled using the `n_sims` argument). Total simulations will be the product of `n_loops` and `n_sims` arguments. If more than one core (`cores`) is specified, then the simulations will be run in parallel to speed up the process. Low numbers of `n_sims` and high numbers of `n_loops` will be easier on RAM, but may be slower. The optimum ratio of `n_sims` to `n_loops` will depend on the amount of RAM and number of cores the user's computer has. In any case, this function simplifies the process of running thousands of simulations of the same survey and the simulated data can then be supplied to survey-based or model-based analyses that require simulation testing.

@@

Processing time will system (i.e. amount of RAM and number of cores) and setting (i.e. `n_loops` and `n_sims` ratio) dependent. The `test_survey` function will print a progress bar, generated using the progress package [@csardi2016], which details percent completion and will also include an eta after the first step of the loop completes. Keep in mind that thousands of simulations of hundreds of survey designs may take days to run. The `test_surveys` function therefore includes an option for exporting intermediate results to a local directory, via the `export_dir` argument, and the `resume_test` function can be used resume a `test_surveys` run that had to be stopped part way through the process. The final object produced will be a list that includes all objects from `sim_abundance` and `sim_distribution` with the table of survey designs tested (named `surveys`) and tables produced by `strat_error` that end with the names `_strat_error` and `_strat_error_stats`. These tables include a `survey` column to allow merging of the survey protocol table with the error tables. Objects produced by `sim_survey` (set and sampling details) and `run_strat` (full stratified analysis results) are not retained to minimize object size. Like other core functions, some convenience functions are included in **`SimSurvey`** for creating interactive plots of the results from `test_surveys`. For instance a series of plotting functions ending in `_fan` produces fan charts where stratified estimates of abundance from each simulated survey are converted into a series of quantiles to depict the probability that estimates fall within a particular range. True values of abundance available to the survey are overlaid on the series of probability envelopes. These plots help visually assess the level of precision and bias from a specific set of survey protocol. The three lines of code below will produce interactive fan charts for stratified estimates of total abundance, abundance at length and abundance at age, respectively. A subset of these plots have been combined into static plots (Figure \@ref(fig:fan1), \@ref(fig:fan2), \@ref(fig:fan3)).

@@

```{r rank, fig.cap = "- Divergent dot plot of the percision and accuracy (RMSE) of length based stratified estimates of abundance, and total sampling effort (number of sets [$N_{sets}$] and length measurements [$N_{measured}$]), under various sampling protocol (set density [$D_{sets}$] and maximum number of lengths measured per set [$M_{lengths}$])."}

@@

- Feedback and advice: Aaron Adamack, Jonathan Babyn, Alejandro Buren, Noel Cadigan, Karen Dwyer, Geoff Evans, Brian Healey, Paul Higdon, Danny Ings, Mariano Koen-Alonso, Joanne Morgan, Keith Lewis, Derek Osborne, Pierre Pepin, Dwayne Pittman, Don Power, Craig Purchase, Greg Robertson, Martha Robertson, Mark Simpson, Brad Squires, Don Stansbury, Peter Upward
- Package contributions: Jonathan Babyn, Greg Robertson
- Dave Cote, Keith Lewis and Greg Robertson for a review of a previous version of this manuscript
- NSERC visiting-fellow program

@@

>>>>

**1\.** Fisheries scientists are often faced with two critical questions: (i) how many samples are needed to adequately represent a population? And, (ii) how should samples from multi-stage sampling be analyzed?

**2\.** For assessments based on fisheries-independent surveys, our ability to estimate population characteristics, such as abundance at age, is affected by multiple levels of sampling (i.e. the number of sets, lengths and ages sampled in a survey) as well as the methods used to analyze the data.

**3\.** Though simulations are a useful tool for exploring the efficacy of specific sampling strategies and statistical methods, system and survey complexity have limited the number of frameworks that simulation test the full sampling and analytical pathway of fisheries-independent survey data.

**4\.** Here we introduce the `R` package **`SimSurvey`**, which has been designed to simplify the process of simulating surveys of spatial fish populations. The packages includes methods for simulating age-structured populations that varies in space and time as well as multi-stage surveys that sample the virtual population. A series of parameters can be used to control the characteristics of the population and the sampling intensity of the survey. **`SimSurvey`** also include a function for running a stratified-random analysis of the simulated data.

**5\.** **`SimSurvey`** can serve as a convenient, accessible and flexible platform for simulating a wide range of sampling strategies of a range of species. Various design- or model-based approaches can be applied to the results to test the efficacy of different approaches.

**Key-words:** length and age distribution; simulation; spatial correlation; stratified analysis; survey design

@@

It is well known that the accuracy and reliability of results from any population monitoring program hinges on good survey design and analysis [@thomas2010]. Ideally, the design and analysis of large and costly survey should be simulation tested before embarking on expensive field programs. This is especially true for the large-scale and costly fisheries-independent trawl surveys conducted by numerous regional fisheries management organizations. These surveys have become a mainstay in the management of dynamic fish stocks as they provide indices of population abundance as well as estimates of various population characteristics such as length and age frequencies [@Pennington1998]. This information forms the basis of many stock assessments throughout the world. Efficient, effective and informed management decisions therefore require surveys and analyses that maximize information while minimizing the expense of data collection. While simulations provide a platform for exploring solutions to this optimization problem, building the necessary simulation framework is not a trivial task given the multi-stage nature of the typical sampling program and the complexity of the population processes we aim to represent. The complexity of this problem is perhaps why simulations that test the full sampling and analytical pathway of fisheries-independent surveys are rare [but see @puerta2019; @schnute2003].

Here we document **`SimSurvey`**, an `R` package designed to facilitate realistic simulations of fisheries-independent trawl surveys. This simulation framework has similarities to those presented by Schnute and Haigh [-@schnute2003] and Puerta et al. [-@puerta2019], however, efforts were focused on developing a series of general and accessible functions to simplify the process of testing multiple sampling scenarios and analytical pathways. The steps taken to simulate surveys of spatial, age-structure populations are outlined below, with specific emphasis of a case study that motivated this work.

@@

The **`SimSurvey`** package was written in the programming language `R` [@R] and, in short, the package allows for the simulation of random or stratified-random surveys of an age-structured population that varies in space and time. **`SimSurvey`** relies heavily on functions from the **`data.table`** [@dowle2017], **`raster`** [@hijmans2016] and **`plotly`** [@sievert2018] packages for their efficient data processing, geographic and plotting facilities, respectively. All the source `R` code behind **`SimSurvey`** is available on GitHub (https://github.com/PaulRegular/SimSurvey) and the package can be installed via GitHub:

@@

```{r grid, fig.cap = "- Plots produced by `plot_grid` when supplied an raster object produced by `make_grid` a) using default case study settings and b) settings generates a grid with four divisions and a more linear depth gradient. In these plots, the color gradient represents depth, the grey lines delineate divisions and white lines delineate strata."}

@@

As noted above, the amount of RAM a user's computer has may limit the utility of the `sim_survey` function for running thousands of simulations of the same survey. The `sim_survey_parallel` function was therefore constructed to facilitate the running of thousands of simulations without running into RAM limitations. This function is set-up to run multiple `sim_survey` calls in parallel using the **`doParallel`** package [@weston2015] and, as such, multiple loops can be run using the `n_loops` argument and within each loop multiple simulations can be run (controlled using the `n_sims` argument). Total simulations will be the product of `n_loops` and `n_sims` arguments. If more than one core (`cores`) is specified, then the simulations will be run in parallel to speed up the process. Low numbers of `n_sims` and high numbers of `n_loops` will be easier on RAM, but may be slower. The optimum ratio of `n_sims` to `n_loops` will depend on the amount of RAM and number of cores the user's computer has. In any case, this function simplifies the process of running thousands of simulations of the same survey and the simulated data can then be supplied to survey-based or model-based analyses that require simulation testing.

@@

Processing time will system (i.e. amount of RAM and number of cores) and setting (i.e. `n_loops` and `n_sims` ratio) dependent. The `test_survey` function will print a progress bar, generated using the **`progress`** package [@csardi2016], which details percent completion and will also include an eta after the first step of the loop completes. Keep in mind that thousands of simulations of hundreds of survey designs may take days to run. The `test_surveys` function therefore includes an option for exporting intermediate results to a local directory, via the `export_dir` argument, and the `resume_test` function can be used resume a `test_surveys` run that had to be stopped part way through the process. The final object produced will be a list that includes all objects from `sim_abundance` and `sim_distribution` with the table of survey designs tested (named `surveys`) and tables produced by `strat_error` that end with the names `_strat_error` and `_strat_error_stats`. These tables include a `survey` column to allow merging of the survey protocol table with the error tables. Objects produced by `sim_survey` (set and sampling details) and `run_strat` (full stratified analysis results) are not retained to minimize object size. Like other core functions, some convenience functions are included in **`SimSurvey`** for creating interactive plots of the results from `test_surveys`. For instance a series of plotting functions ending in `_fan` produces fan charts where stratified estimates of abundance from each simulated survey are converted into a series of quantiles to depict the probability that estimates fall within a particular range. True values of abundance available to the survey are overlaid on the series of probability envelopes. These plots help visually assess the level of precision and bias from a specific set of survey protocol. The three lines of code below will produce interactive fan charts for stratified estimates of total abundance, abundance at length and abundance at age, respectively. A subset of these plots have been combined into static plots (Figure \@ref(fig:fan1), \@ref(fig:fan2), \@ref(fig:fan3)).

@@

```{r rank, fig.cap = "- Divergent dot plot of the precision and accuracy (RMSE) of length based stratified estimates of abundance, and total sampling effort (number of sets [$N_{sets}$] and length measurements [$N_{measured}$]), under various sampling protocol (set density [$D_{sets}$] and maximum number of lengths measured per set [$M_{lengths}$])."}

@@

**Feedback and advice:** Aaron Adamack, Jonathan Babyn, Alejandro Buren, Noel Cadigan, Karen Dwyer, Geoff Evans, Brian Healey, Paul Higdon, Danny Ings, Mariano Koen-Alonso, Joanne Morgan, Keith Lewis, Derek Osborne, Pierre Pepin, Dwayne Pittman, Don Power, Craig Purchase, Greg Robertson, Martha Robertson, Mark Simpson, Brad Squires, Don Stansbury, Peter Upward

**Package contributions:** Jonathan Babyn, Greg Robertson

**Review of previous draft:** Dave Cote, Keith Lewis, Greg Robertson

**Funding:** NSERC visiting-fellow program

@@

<!-- # Quotes -->

<!-- "To effectively manage rare populations, accurate monitoring data are critical. Yet many monitoring programs are initiated without careful consideration of whether chosen sampling designs will provide accurate estimates of population parameters. Obtaining accurate estimates is especially difficult when natural variability is high, or limited budgets determine that only a small fraction of the population can be sampled." - Morrison et al. (2008) Popul Ecol 50:417–425 -->

<!-- "Fishery independent surveys provide valuable information on fish population characteristics and play important roles in fisheries stock assessment (Gunderson, 1993; Rago, 2005). The importance of fishery independent surveys has been well recognized, but the important role of sampling design for these surveys is less appreciated. Often, traditional survey methods are used without considering survey designs that maximize the precision and accuracy of the estimator (Mier and Picquelle, 2008). Fishery independent surveys are usually expensive and time-consuming, and an efficient sampling design is key to the success of these surveys. Recently, investigators have begun to pay more attention to the importance of survey sampling and are exploring a variety of aspects of designs (Bez, 2002; Brown, 2003; Kimura and Somerton, 2006). Based on the previous studies, we compared the performance of traditional and adaptive sampling designs with emphasis on simulation application for performance comparison." - Yu et al. (2012) Fisheries Research 113: 173–181 -->

<!-- # Examples -->

<!-- Good examples of papers on packages: -->
<!-- - https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0092725#s4 -->
<!-- - file:///C:/Users/RegularP/Downloads/Mildenberger_et_al-2017-Methods_in_Ecology_and_Evolution.pdf -->

<!-- # READ -->

<!-- - https://onlinelibrary.wiley.com/doi/pdf/10.1046/j.1467-2960.2001.00047.x -->
